Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-07-01 overall.tex
introduce  \ifreview
  to toggle between reviewer comments/not, so as to make it easier to get a handle on
  net space used.
Address referee's comments and marginal space saving

Comments to address in this section:
\begin{enumerate}
    \item REVIEWER 2: In Figure 4, there are some inconsistencies. The figure
    caption shows there are only 62175 measures, but in the scatter plot
    (right) there are 62555 data points. In addition, according to the
    distribution of the number of work nodes (left), it does not look like
    the mean number of work available is 691 as the figure caption
    illustrates.
    \item REVIEWER 3: t’s interesting that Feb/March have such an increase in
    backfill availability in 2017. The authors attribute this to hardware
    upgrades for the DTNs. It’s mentioned that the load is only 3.6\% now and
    therefore further upgrades will likely not cause similar improvement.
    What was the previous CPU load? Why were particularly low percentages
    utilized in April/May 2016?
    \item REVIEWER 4: “Genat4” at the end of Section 4 is a typo.
    \item REVIEWER 4: Please explain why you used Titan at all, in spite of
    your statement that it is 1/2 to 1/3 as efficient as other resources. Was
    it merely that you were given so much time on this resource that the
    inefficiency doesn’t matter?
    \item REVIEWER 4: Why didn’t you benchmark the performance of your
    workload on other HPC resources that might have delivered better
    performance, prior to reengineering your entire workflow to deal with the
    challenges posed by Titan, a machine that doesn’t seem to justify the
    pain?
    \item REVIEWER 4: The INCITE program operational requirements for the
    LCFs are 90\% uptime. Using large batches of Geant4 jobs that run on a
    single node or less to reach this metric sounds like gaming the system
    rather than an accomplishment.
    \item REVIEWER 5: In Figure 3, it shows the unused core hours are much
    higher than those used by ATLAS. What are the reasons for such low usage?
\end{enumerate}

Currently, ATLAS deploys 20 instances of the PanDA Broker on 4 Titan's DTNs,
5 instances per DTN\@. Each broker submits and manages the execution of 15 to
300 ATLAS jobs, one job for each Titan's worker node, and a theoretical
maximum concurrent use of 96,000 cores. Since November 2015, PanDA Brokers
have operated only in backfill mode, without a defined time allocation, and
running at the lowest priority on Titan.

We evaluate the efficiency, scalability and reliability of the deployment of
PanDA WMS on Titan by characterizing the behavior of both PanDA Broker and
AthenaMP\@. We discuss challenges and limitations of our approach at multiple
levels arising from the specifics of workload, middleware and methods. All
the measurements were performed between January 2016 and February 2017,
hereafter called ‘experiment time window

@@

Fig.~\ref{fig:backfill-utilization} shows the total backfill availability on
Titan (gray bars) and the number of core-hours of that availability used by
ATLAS (blue bars) during the experiment time window. ATLAS consumed a total
of 51.4M core-hours, for an average of $\approx$3.7M core-hours a month, with
a minimum of 1.7M core-hours in April 2016 and a maximum 7.9M core-hours in
February 2017.

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and
    CPU core-hours used by ATLAS (blue). GPU core-hours unaccounted for as
    they cannot be used by ATLAS. Efficiency of PanDA Brokers defined as
    percentage of total Titan's backfill availability used by ATLAS (Red
    labels).}
\label{fig:backfill-utilization}
\end{figure}

@@

On February 2017, PanDA Brokers used almost twice as much total backfill
availability than in any other month (preliminary results for March 2017
displayed in Fig.~\ref{fig:backfill-utilization} confirm this trend). No
relevant code update was made during that period and logs indicated that the
brokers were able to perform faster. This is likely due to hardware upgrades
on the DTNs. The absence of continuous monitoring of those nodes does not
allow to quantify bottlenecks but spot measurements of their load indicate
that a faster CPU and better networking were likely responsible for the
improved performance. Investigations showed an average CPU load of 3.6\% on
the upgraded DTNs. As such, further hardware upgrades seem unlikely to
improve significantly the performance of PanDA Brokers.

Every detector simulation executed on Titan process 100 events. This number
of events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events simulated but not all events take the same
time to be simulated. 1 event simulation takes from $\approx$2 to $\approx$40
minutes, with a mean of $\approx$14 minutes. Considering that each worker
node process up to 16 events concurrently, 100 events takes an average of 105
minutes to process. As such, PanDA brokers do not use backfill availability
with less than 105 minutes walltime.

Fig.~\ref{fig:backfill-distrib} shows the distribution of backfill
availability on Titan as a function of number of nodes and the time of their
availability (i.e., walltime). We recorded these data by polling Titan's Moab
scheduler at regular intervals during the experiment window time. The mean
number of nodes was 691, and their mean walltime was 126 minutes. Detector
simulations of 100 events, enable to use down to 5/6 of the mean walltime of
backfill availability. As such, it offers a good compromise for PanDA Broker
efficiency.

\begin{figure*}[!t]
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_wnodes_distribution.pdf}
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_walltime_distribution.pdf}
    \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_avail.png}
   \vspace{-0.1in}
    \caption{Backfill availability: distribution of number of work nodes
    (left) and of walltime in minutes (center). Scatter plot of the two
    variables (right). 62175 measures: mean number of work nodes available
    691; mean walltime available 126 minutes.}

\label{fig:backfill-distrib}
\end{figure*}

@@

In September 2016, increasing the number of concurrent PanDA brokers from 4
to 20 markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization})
but further research is undergoing to understand whether an even larger
number of brokers would yield similar results. This research focuses on
evaluating the overheads of input/output files staging, including its impact
on DTNs, and on an alternative design of PanDA Broker that enables the
concurrent submission of multiple MPI scripts~\cite{barreiro2016panda}.

The current design and architecture of the PanDA Broker is proving to be as
reliable as PanDA Pilot when used on the WLCG. Between Jan 2016 and Feb 2017,
the overall failure rate of all the ATLAS detector simulation jobs was 14\%,
while the failure rate of jobs submitted to Titan was a comparable 13.6\%.
PanDA Brokers were responsible for around the 19\% of the failures, compared
to the 29\% of failures produced by the JobDispatcher module of the PanDA
Server, and the 13\% failures produced by the Geant4 toolkit. The current
failure rate of the PanDA Brokers confirms the benefits of reusing most of
the code base of the PanDA Pilot for implementing the PanDA Broker. It also
shows that adopting third-party libraries like RADICAL-SAGA did not have a
measurable adverse effect on reliability.

% -----------------------------------------------------------------------------
\subsection{Characterizing the Detector Simulation on Titan}
\label{ssec:athenamp_titan}

We use two main parameters to measure the performance of the detector
simulation jobs submitted to Titan: (i) the time taken to setup AthenaMP; and
(ii) the distribution of the time taken by the Geant4 toolkit to simulate a
certain number of events.

AthenaMP has an initialization and configuration stage. At initialization
time, AthenaMP is assembled from a large number of shared libraries,
depending on the type of payload that will have to be computed. Once
initialized, every algorithm and service of AthenaMP is configured by a set
of Python scripts. Both these operations result in a large number of read
operations on the filesystem shared between the worker nodes and the DTNs,
including the operations required to access small python scripts.

Initially, all the shared libraries of AthenaMP and the python scripts for
its configuration were stored on the Spider 2 Lustre file system. However,
the I/O patterns of the initialization and configuration stages degraded the
performance of the filesystem. This was addressed by moving the AthenaMP
distribution to a read-only NFS directory, shared among DTNs and worker
nodes. NFS eliminated the problem of metadata contention, improving metadata
read performance from $\approx$6,300 seconds on Lustre to $\approx$1,500
seconds on NFS.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}
   \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when
    placing 2 simulations (h1) or 1 simulation (h2) per Titan's CPU. 2
    simulation use 16 cores per node, 1 simulation 8. 50 Events; 1 Titan
    worker nodes; 16 work threads per node; 100 events per node.}

@@

The differences in performance between Titan and the BNL and SIGNET Grid
sites are due to the FP scheduler competition and the availability of newer
processors. The CPUs at the Grid sites have one FP scheduler per core and are
on average newer than the CPU of Titan. The heterogeneity of the Grid sites'
CPUs explain the higher performance variance we observed compared to the
performance consistency we measured on Titan.

@@

Overall, the average time taken to read events from input files stored on
Lustre is 1,320, comparable to the time taken to read the file required by
assembling AthenaMP from NFS. Preliminary investigation shows that this time
could be reduced to 40 seconds by loading the event descriptions into the RAM
disk available on each worker node. Events descriptions could be transferred
from Lustre to the RAM disk while configuring and initializing AthenaMP,
almost halving the time currently required by initiating a Genat4 simulation.

>>>>

\ifreview
Comments to address in this section:
\begin{enumerate}
    % \item REVIEWER 2: In Figure 4, there are some inconsistencies. The
    % figure caption shows there are only 62175 measures, but in the scatter
    % plot (right) there are 62555 data points. In addition, according to the
    % distribution of the number of work nodes (left), it does not look like
    % the mean number of work available is 691 as the figure caption
    % illustrates.
    \item REVIEWER 3: It’s interesting that Feb/March have such an increase
    in backfill availability in 2017. The authors attribute this to hardware
    upgrades for the DTNs. It’s mentioned that the load is only 3.6\% now and
    therefore further upgrades will likely not cause similar improvement.
    What was the previous CPU load? Why were particularly low percentages
    utilized in April/May 2016?
    % \item REVIEWER 4: “Genat4” at the end of Section 4 is a typo.
    % \item REVIEWER 4: Please explain why you used Titan at all, in spite of
    % your statement that it is 1/2 to 1/3 as efficient as other resources.
    % Was it merely that you were given so much time on this resource that
    % the inefficiency doesn’t matter?
    \item REVIEWER 4: Why didn’t you benchmark the performance of your
    workload on other HPC resources that might have delivered better
    performance, prior to reengineering your entire workflow to deal with the
    challenges posed by Titan, a machine that doesn’t seem to justify the
    pain?
    \item REVIEWER 4: The INCITE program operational requirements for the
    LCFs are 90\% uptime. Using large batches of Geant4 jobs that run on a
    single node or less to reach this metric sounds like gaming the system
    rather than an accomplishment.
    % \item REVIEWER 5: In Figure 3, it shows the unused core hours are much
    % higher than those used by ATLAS\@. What are the reasons for such low
    % usage?
\end{enumerate}
\fi

Currently, ATLAS deploys 20 instances of the PanDA Broker on 4 Titan's DTNs,
5 instances per DTN\@. Each broker submits and manages the execution of 15 to
300 ATLAS jobs, one job for each Titan's worker node, and a theoretical
maximum concurrent use of 96,000 cores. Since November 2015, PanDA Brokers
have operated only in backfill mode, without a defined time allocation, and
running at the lowest priority on Titan. Therefore, ATLAS contributed to
increase the Titan's overall utilization without affecting other users.

We evaluate the efficiency, scalability and reliability of the deployment of
PanDA WMS on Titan by characterizing the behavior of both PanDA Broker and
AthenaMP\@. We discuss challenges and limitations of our approach at multiple
levels arising from the specifics of workload, middleware and methods. All
the measurements were performed between January 2016 and February 2017,
hereafter called `experiment time window'.

@@

Fig.~\ref{fig:backfill-utilization} shows the total backfill availability on
Titan (gray bars) and the number of core-hours of that availability used by
ATLAS (blue bars) during the experiment time window. ATLAS consumed a total
of 51.4M core-hours, for an average of $\approx$3.7M core-hours a month. 
% , with a minimum of 1.7M core-hours in April 2016 and a maximum 7.9M
% core-hours in February 2017.

\begin{figure}[!t]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \vspace{-0.3in}
    \caption{Titan's total backfill availability: CPU core-hours (gray) and
    CPU core-hours used by ATLAS (blue). GPU core-hours unaccounted for as
    they cannot be used by ATLAS\@. Efficiency of PanDA Brokers defined as
    percentage of total Titan's backfill availability used by ATLAS (Red
    labels).}
\label{fig:backfill-utilization}
\end{figure}

@@

On February 2017, PanDA Brokers used almost twice as much total backfill
availability than in any other month (preliminary results for March 2017
displayed in Fig.~\ref{fig:backfill-utilization} confirm this trend). No
relevant code update was made during that period and logs indicated that the
brokers were able to perform faster. This is likely due to hardware upgrades
on the DTNs. The absence of continuous monitoring of those nodes does not
allow to quantify bottlenecks but spot measurements of their load indicate
that a faster CPU and better networking were likely responsible for the
improved performance. Investigations showed an average CPU load of 3.6\% on
the upgraded DTNs. As such, further hardware upgrades seem unlikely to
improve significantly the performance of PanDA Brokers.\mtnote{Asked Segey
and Danila for values previous HW upgrade}

Every detector simulation executed on Titan process 100 events. This number
of events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events simulated but not all events take the same
time to be simulated. One event simulation takes from $\approx$2 to
$\approx$40 minutes, with a mean of $\approx$14 minutes. Considering that
each worker node process up to 16 events concurrently, 100 events takes an
average of 105 minutes to process. As such, PanDA brokers do not use backfill
availability with less than 105 minutes walltime.

Fig.~\ref{fig:backfill-distrib} shows backfill availability on Titan as a
function of number of nodes and the time of their availability (i.e.,
walltime). We recorded these data by polling Titan's Moab scheduler at
regular intervals during the experiment window time. The mean number of nodes
was 691, and their mean walltime was 126 minutes. Detector simulations of 100
events, enable to use down to 5/6 of the mean walltime of backfill
availability. As such, it offers a good compromise for PanDA Broker
efficiency.

%\begin{figure*}[!t]
\begin{figure}[!t]
    % \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_wnodes_distribution.pdf}
    % \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_walltime_distribution.pdf}
    % \includegraphics[clip,width=0.32\textwidth]{figures/titan_backfill_avail.png}
    \includegraphics[clip,width=\columnwidth]{figures/titan_backfill_avail.png}
   \vspace{-0.1in}
    \caption{
    % Backfill availability: distribution of number of work nodes (left) and
    % of walltime in minutes (center). Scatter plot of the two variables.
    % (right).
    62555 measures of Backfill availability on Titan during the experiment
    time window. Mean number of work nodes available 691; mean walltime
    available 126 minutes.}
\label{fig:backfill-distrib}
%\end{figure*}
\end{figure}

@@

In September 2016, increasing the number of concurrent PanDA brokers from 4
to 20 markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization})
but further research is undergoing to understand whether an even larger
number of brokers would yield similar results. This research focuses on
evaluating the overheads of input/output files staging, including its impact
on DTNs, and on an alternative design of PanDA Broker that enables the
concurrent submission of multiple MPI scripts~\cite{barreiro2016panda}. The
results of these investigation will contribute to improve the efficienty of
PanDA Brokers beyond the 30\% limit showed in
Fig.~\ref{fig:backfill-utilization}.

The current design and architecture of the PanDA Broker is proving to be as
reliable as PanDA Pilot when used on the WLCG\@. Between Jan 2016 and Feb
2017, the overall failure rate of all the ATLAS detector simulation jobs was
14\%, while the failure rate of jobs submitted to Titan was a comparable
13.6\%. PanDA Brokers were responsible for around the 19\% of the failures,
compared to the 29\% of failures produced by the JobDispatcher module of the
PanDA Server, and the 13\% failures produced by the Geant4 toolkit. 
% The current failure rate of the PanDA Brokers confirms the benefits of
% reusing most of the code base of the PanDA Pilot for implementing the PanDA
% Broker. It also shows that adopting third-party libraries like RADICAL-SAGA
% did not have a measurable adverse effect on reliability.

% ---------------------------------------------------------------------------
\subsection{Characterizing the Detector Simulation on Titan}
\label{ssec:athenamp_titan}

We use two main parameters to measure the performance of the detector
simulation jobs submitted to Titan: (i) the time taken to setup AthenaMP\@;
and (ii) the distribution of the time taken by the Geant4 toolkit to simulate
a certain number of events.

AthenaMP has an initialization and configuration stage. At initialization
time, AthenaMP is assembled from a set of shared libraries, depending on the
type of payload that will have to be computed. Once initialized, every
algorithm and service of AthenaMP is configured via Python scripts. Both
these operations result in read operations on the filesystem shared between
the worker nodes and the DTNs, including the operations required to access
small python scripts.

Initially, all the shared libraries and the python scripts of AthenaMP were
stored on the OLCF Spider 2 Lustre file system. However, the I/O patterns of
the initialization and configuration stages degraded the performance of the
filesystem. This was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among Titan's DTNs and worker nodes. NFS
eliminated the problem of metadata contention, improving metadata read
performance from $\approx$6,300 seconds on Lustre to $\approx$1,500 seconds
on NFS\@.

@@

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison_vsquashed.pdf}
   \vspace{-0.3in}
    \caption{Distributions of the time taken to simulate one event when
    placing 2 simulations (h1) or 1 simulation (h2) per Titan's CPU\@. 2
    simulation use 16 cores per node, 1 simulation 8. 50 Events; 1 Titan
    worker nodes; 16 work threads per node; 100 events per node.}

@@

The differences in performance between Titan and the BNL and SIGNET Grid
sites are due to the FP scheduler competition and the availability of newer
processors. The CPUs at the Grid sites have one FP scheduler per core and are
on average newer than the CPU of Titan. The heterogeneity of the Grid sites'
CPUs explain the higher performance variance we observed compared to the
performance consistency we measured on Titan. This difference in performance
is compensated by the amount of resources available on Titan (capable of
executing up to 30\%/year of the ATLAS detector simulations) and by the lack
of further resources available to WLCG\@. Also, it should be noted that Titan
is at the end of its life-cycle and that Summit, Titan's successor, will
offer cutting-edge performances.

@@

Overall, the average time taken to read events from input files stored on
Lustre is 1,320, comparable to the time taken to read the file required by
assembling AthenaMP from NFS\@. Preliminary investigation shows that this
time could be reduced to 40 seconds by loading the event descriptions into
the RAM disk available on each worker node. Events descriptions could be
transferred from Lustre to the RAM disk while configuring and initializing
AthenaMP, almost halving the time currently required by initiating a Geant4
simulation.
